---
title: 'COVID-19 Vaccinations Data: Some Visualisations for Italy'
author: 'Luca Baggi'
date: '2021-01-23'
slug: 'covid-19-vaccinations-italy-visualisations'
categories: []
tags:
  - covid19
  - italy
  - visualisation
ShowToc: true
---

```{r chunks, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, 
                      # tidy = "styler", fig.width = 8, fig.height = 5,
                      echo = TRUE, dpi = 300, cache.lazy = FALSE)

library(tidyverse)
theme_set(theme_minimal())
```

Now that we have wrangled the data a bit, we can proceed with some visualisations. We want to plot three things:

-   How many vaccinations are administered daily.
-   How many doses have been administered so far and their ratio.
-   See how regions perform in terms of doses administered and doses received.

# Data Wrangling

Let's load the data:

```{r data, message=FALSE}
read_csv(
  'https://raw.githubusercontent.com/orizzontipolitici/covid19-vaccine-data/main/data_ita/doses_by_date_ita.csv',
) -> doses_by_date 

read_csv(
  'https://raw.githubusercontent.com/orizzontipolitici/covid19-vaccine-data/main/data_ita/vaccinations_by_area_ita.csv', col_types = cols(area = col_factor()) ) -> vaccinations_by_area
```

These two datasets are incompatible: first, data grouped by area needs to be grouped to the Italian level. Then, we need to transform deliveries data into the long format.

Let's start with grouping vaccination data. Columns beyond the twelfth are not needed: basically, they are the sum of the doses injected each day... which is exactly what we will be computing within `summarise()`!

```{r vaccinations_by_date, message=FALSE}
vaccinations_by_area %>%
  # the other columns are the same as the summarised one
  select(1:12) %>%
  group_by(data, fornitore) %>%
  summarise(across(where(is.numeric), sum)) %>%
  ungroup() -> vaccinations_by_date
```

Then, let's move to vaccine data. This seems a bit more challenging at first, as we need to pack the two columns with the total deliveries of each supplier into a single one. `pivot_longer()` comes to rescue!

```{r doses_long}
doses_by_date %>%
  # we will just need data and the two supplier cols
  select(-totale_dosi_consegnate, -totale_dosi) %>%
  # to get prettier names into the new column
  rename(
    'Pfizer/BioNTech' = totale_pfizer,
    Moderna = totale_moderna
  ) %>%
  # pivot longer magic:
  pivot_longer(c('Pfizer/BioNTech', Moderna),
               # will have values Pfizer/BioNTech and Moderna
               names_to = 'fornitore',
               # will report the deliveries of the day
               values_to = 'consegne') %>%
  # we need to add cumulative doses available at each date
  group_by(fornitore) %>%
  mutate(dosi_totali = cumsum(consegne)) -> doses_by_date_long
```

# Vaccinations: First and Second Doses

Now that the two are compatible, we could join them. However, given the structure of the data, we'd better create two new separate tables: for instance, since we cannot observe to who the first and second shots are administered, we can create a new table with this info only.

```{r}
vaccinations_by_date %>%
  select(data, fornitore, prima_dose, seconda_dose) %>%
  pivot_longer(cols = !c(data, fornitore),
               names_to = 'dose',
               values_to = 'count') %>%
  group_by(fornitore, dose) %>%
  mutate(
    total_count = cumsum(count)
  ) -> vaccinations_long
```

Vaccines from Moderna are so few right now that we might as well group by it. In the future, comparing how shots are administered by each supplier would bear greater value.

```{r, message=FALSE}
vaccinations_long %>%
  group_by(data, dose) %>%
  summarise(across(where(is.numeric), sum)) %>%
  # swap the order of the factor
  mutate(dose = fct_rev(dose)) %>%
  ggplot(aes(data, total_count, fill = dose)) +
  geom_area() +
  scale_fill_viridis_d(begin = 0.95, end = 0.55) +
  labs(
    title = 'Vaccinations in Italy',
    subtitle = 'Cumulative First and Second Doses Administered',
    x = NULL,
    y = NULL,
    fill = NULL
  ) +
  theme(legend.position = 'bottom')
```
We can also show it as a percentage: it's a bit more complicated since we have to move from wider to create the new percentage columns and then back to longer for plotting.

```{r, message=FALSE}
vaccinations_long %>%
  group_by(data, dose) %>%
  summarise(across(where(is.numeric), sum)) %>%
  # pivot back to wider to create new columns
  pivot_wider(
    !total_count,
    names_from = dose,
    values_from = c(count)
  ) %>%
  # take out redundant text from column names
  rename_with( ~ stringr::str_remove(.x, c('count_'))) %>%
  mutate(across(
    # apply it to these cols  
    c(prima_dose, seconda_dose),
    # replace with
    ~ .x / (prima_dose + seconda_dose)
  )) %>%
  # pivot to longer
  pivot_longer(
    # don't use data
    !data,
    # names and values
    names_to = 'dose',
    values_to = 'pct'
  ) %>%
  ggplot(aes(data, pct, fill = dose)) +
  geom_area() +
  scale_fill_viridis_d(begin = 0.55, end = 0.95) +
  labs(
    title = 'Vaccinations in Italy',
    subtitle = 'Cumulative First and Second Doses Administered',
    x = NULL,
    y = NULL,
    fill = NULL
  ) +
  scale_y_continuous(labels = scales::percent) +
  theme(legend.position = 'bottom')
```

## Daily Vaccinations Data

We can also plot a line with daily shots (again, data is grouped by `fornitore`):

```{r, message=FALSE}
vaccinations_long %>%
  group_by(data, dose) %>%
  summarise(across(where(is.numeric), sum)) %>%
  ungroup() %>%
  ggplot(aes(data, count, col = dose)) +
  geom_line() +
  # `begin` and `end` are swapped as factors are not swapped as in the graph above
  scale_color_viridis_d(begin = 0.55, end = 0.95) +
  labs(
    title = 'Vaccinations in Italy',
    subtitle = 'First and Second Doses Administered',
    x = NULL,
    y = NULL,
    color = NULL
  ) +
  theme(legend.position = 'bottom')
```

## Putting these together

We will be using a call to `ggpubr::ggarrange()`:

```{r, echo=FALSE, message=FALSE}
p1 <- vaccinations_long %>%
  group_by(data, dose) %>%
  summarise(across(where(is.numeric), sum)) %>%
  # swap the order of the factor
  mutate(dose = fct_rev(dose)) %>%
  ggplot(aes(data, total_count, fill = dose)) +
  geom_area() +
  scale_fill_viridis_d(begin = 0.95, end = 0.55) +
  labs(
    title = 'Vaccinations in Italy',
    subtitle = 'Total doses injected, by number of shot',
    x = NULL,
    y = NULL,
    fill = NULL
  ) +
  theme(legend.position = 'none')

p2 <- vaccinations_long %>%
  group_by(data, dose) %>%
  summarise(across(where(is.numeric), sum)) %>%
  # pivot back to wider to create new columns
  pivot_wider(
    !total_count,
    names_from = dose,
    values_from = c(count)
  ) %>%
  # take out redundant text from column names
  rename_with( ~ stringr::str_remove(.x, c('count_'))) %>%
  mutate(across(
    # apply it to these cols  
    c(prima_dose, seconda_dose),
    # replace with
    ~ .x / (prima_dose + seconda_dose)
  )) %>%
  # pivot to longer
  pivot_longer(
    # don't use data
    !data,
    # names and values
    names_to = 'dose',
    values_to = 'pct'
  ) %>%
  ggplot(aes(data, pct, fill = dose)) +
  geom_area() +
  scale_fill_viridis_d(begin = 0.55, end = 0.95) +
  labs(
    title = NULL,
    subtitle = 'Percentage of first and second shot injected, each day',
    x = NULL,
    y = NULL,
    fill = NULL
  ) +
  scale_y_continuous(labels = scales::percent) +
  theme(legend.position = 'bottom')

ggpubr::ggarrange(p1, p2, nrow = 2)
```

# Comparing First and Second Doses

```{r, message=FALSE, warning=FALSE}
vaccinations_by_date %>%
  filter(fornitore != 'Moderna') %>%
  select(data, prima_dose, seconda_dose) %>%
  mutate(lead = lead(seconda_dose, 20)) %>%
  ggplot(aes(data, cumsum(prima_dose))) +
  geom_line(col = 'salmon') +
  geom_line(aes(y = cumsum(lead)), col = 'cornflowerblue') +
  labs(
    title = 'Total First and Second Doses Injected',
    subtitle = 'Red is the total number of first doses\nBlue is the total number of second doses',
    x = NULL,
    y = NULL
  )
```

# To whom doses are administered to?

Let's use vaccinations data to see how shots are distributed.

## Male and Female

### National Level

```{r}
vaccinations_by_area %>%
  group_by(data) %>%
  summarise(across(where(is.numeric), sum)) %>%
  select(data, sesso_maschile, sesso_femminile) %>%
  pivot_longer(!data, names_to = 'sesso', values_to = 'count') %>%
  ggplot(aes(data, count, fill = sesso)) +
  geom_col() +
  labs(
    title = 'Vaccinations by Sex in Italy',
    subtitle = 'Blue is male, red is female',
    x = NULL,
    y = NULL,
    fill = NULL
  ) +
  theme(legend.position = 'none')
```

Let's see it as percentage:

```{r}
vaccinations_by_area %>%
  # we are not interested in the supplier 
  group_by(data) %>%
  summarise(across(where(is.numeric), sum)) %>%
  select(data, sesso_maschile, sesso_femminile) %>% 
  mutate(
    # transform into percentage
    across(c(sesso_femminile, sesso_maschile), ~ .x / (sesso_maschile + sesso_femminile)),
    # replace NAs with 0
    across(where(is.numeric), ~ coalesce(.x, 0L))
  ) %>%
  pivot_longer(!data, names_to = 'sesso', values_to = 'count') %>%
  ggplot(aes(data, count, fill = sesso)) +
  geom_col() +
  labs(
    title = 'Vaccinations by Sex, per Area',
    subtitle = 'Blue is male, red is female',
    x = NULL,
    y = NULL,
    fill = NULL
  ) +
  scale_y_continuous(labels = scales::percent) +
  theme(legend.position = 'none')
```

At the national level, more women are being vaccinated than men. This may also be due to more of them being health workers.

### Regional Level with `facet_grid()`

We exploit the extra `area` information to see these proportions at the local level. Given the high number of regions, it's won't look pretty, at all! I shall get back to this with `shiny`, soon.

```{r}
vaccinations_by_area %>%
  select(data, area, sesso_maschile, sesso_femminile) %>%
  pivot_longer(!c(data, area), names_to = 'sesso', values_to = 'count') %>%
  ggplot(aes(data, count, fill = sesso)) +
  geom_col() +
  facet_wrap(~ area, nrow = 3) +
  labs(
    title = 'Vaccinations by Sex, per Area',
    subtitle = 'Blue is male, red is female',
    x = NULL,
    y = NULL,
    fill = NULL
  ) +
  # to make the dates fit! thanks to `r-graphics.org`!
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  theme(legend.position = 'none')
```

We did manage to avoid getting an absolute awful graph, though!

## By Category

```{r}
vaccinations_by_area %>%
  # show off some other selection syntax
  select(data, area, operatori_sanitari:ospiti_rsa) %>%
  # for now, no area
  group_by(data) %>%
  summarise(across(where(is.numeric), sum)) %>%
  pivot_longer(
    !data,
    names_to = 'categoria',
    values_to = 'count'
  ) %>%
  ggplot(aes(data, count, fill = categoria)) +
  geom_col() +
  theme(legend.position = 'bottom') +
  scale_fill_viridis_d(begin = 0.2, end = 0.75) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(
    title = 'Vaccinations by category',
    x = NULL,
    y = NULL,
    fill = NULL
  )
```

# Who's doing better?

Let's use some other data to find the region which is performing better. Let's load the data:

```{r, message=F}
read_csv('https://raw.githubusercontent.com/orizzontipolitici/covid19-vaccine-data/main/data_ita/totals_by_area_ita.csv') %>%
  select(-NUTS2, -nome) %>%
  mutate(area = as.factor(area)) -> totals_by_area
```

And then make a plot:

```{r}
totals_by_area %>%
  filter(area != 'ITA') %>%
  ggplot(aes(
    x = dosi_ogni_mille,
    y = vaccinati_ogni_mille
  )) +
  geom_vline(xintercept = mean(totals_by_area$dosi_ogni_mille), color = 'grey') +
  geom_hline(yintercept = mean(totals_by_area$vaccinati_ogni_mille), color = 'grey') +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = area)) +
  # xlim(0,100) +
  # ylim(0,100) +
  labs(
    title = 'Which Region is Ahead?',
    subtitle = 'Doses delivered compared to doses injected',
    x = NULL,
    y = NULL
  )
```
