---
title: Dealing With Implicit Missing Data
author: Luca Baggi
date: '2021-01-28'
slug: dealing-with-implicit-missing-data
categories:
tags:
  - covid19
  - italy
  - visualisation
  - wrangling
ShowToc: True
---

Today I want to test some ways to deal with implicit missing values: namely, creating grids with several commands and performing full joins on our data. Let's use again COVID-19 vaccinations data in Italy, available from the official [repo](https://github.com/italia/covid19-opendata-vaccini/blob/master/dati/somministrazioni-vaccini-latest.csv).

```{r chunk-options, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, 
                      # tidy = "styler", fig.width = 8, fig.height = 5,
                      echo = TRUE, dpi = 300, cache.lazy = FALSE)

library(tidyverse)
```

# Load Data

```{r load-data}
url_vaccinations <-
  'https://raw.githubusercontent.com/italia/covid19-opendata-vaccini/master/dati/somministrazioni-vaccini-latest.csv'

read_csv(url_vaccinations,
         col_types = cols(
           # parse as dates
           data_somministrazione = "D",
           # parse as factors
           fornitore = "f",
           area = "f",
           fascia_anagrafica = "f"
           # the rest, let it be guessed
         )) %>%
  # remove 'categoria' from several column names
  rename_with( ~ stringr::str_remove(.x, 'categoria_')) %>%
  # shorten some other variable names
  rename(
    operatori_sanitari = operatori_sanitari_sociosanitari,
    data = data_somministrazione
  ) %>%
  # create a new column with total vaccinations
  mutate(nuovi_vaccinati = sesso_maschile + sesso_femminile) %>%
  # reorder columns
  relocate(nuovi_vaccinati, .after = 'fascia_anagrafica') ->
  vaccinations

vaccinations %>% head(9)
```

# Create Grids

To fill implicit missing values, we need to check against every possible combinations of factors and dates in our data. This means creating a grid and then performing a full join.

```{r}
expand_grid(
    data = seq.Date(from = min(vaccinations$data),
                    to = max(vaccinations$data),
                    by = 'day'),
    area = forcats::fct_unique(vaccinations$area),
    fornitore = forcats::fct_unique(vaccinations$fornitore),
    fascia_anagrafica = forcats::fct_unique(vaccinations$fascia_anagrafica),
  ) %>% head()
```

The same can be achieved in this way:

```{r}
vaccinations %>%
  expand(
    # create a full sequence between the first and last date
    full_seq(data, 1),
    area, fornitore, fascia_anagrafica)
```

Much neater, uh? Still, there are very many values and once the data gets larger it will take some more time.

```{r}
vaccinations %>%
  full_join(vaccinations %>% expand(data, area, fornitore, fascia_anagrafica),
            by = c('data', 'area', 'fornitore', 'fascia_anagrafica')) %>%
  # sort data
  arrange(area, data) %>%
  # replace NAs that popped up  
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>%
  # don't know why it does not work
  mutate(fascia_anagrafica = fct_inorder(fascia_anagrafica)) -> vaccinations_ita

vaccinations_ita %>% head(10)
```

This is a very great deal of observations, and as of this day there are more missing values than complete ones. This is because Moderna started deliveries more than two weeks later than Pfizer/BioNTech (as the European Center for Disease Prevention and Control approved the vaccine later) and there were also some days between the start of the campaign and approximately January 10th where no vaccines administered.

Now we want to create some smaller datasets to be able to streamline visualisations later.

# Group Data by Age Range

Let's group the data by `data`, `fornitore` and `fascia_anagrafica` to see how vaccinations proceed within the same population range. Furthermore, compute the cumulative totals for each numeric variable.

```{r}
vaccinations_ita %>%
  group_by(data, fascia_anagrafica, fornitore) %>%
  summarise(across(where(is.numeric), sum)) %>%
  mutate(across(where(is.numeric), list(totale = ~ cumsum(.x)))) %>%
  arrange(data, fornitore) %>%
  rename(vaccinati_totale = nuovi_vaccinati_totale) -> vaccinations_by_age_ita

vaccinations_by_age_ita %>% head()
```

## What to do with this?

Before moving on to grouping the data, let's see how many people by age range received a shot (and from which supplier) by age range over time.

```{r}
vaccinations_by_age_ita %>%
  ggplot(aes(data, nuovi_vaccinati, fill = fornitore)) +
  geom_col() +
  facet_wrap(~ fascia_anagrafica) +
  scale_fill_viridis_d(begin = 0.75, end = 0.25) +
  theme_minimal()
```

We can also obtain the totals by age range:

```{r}
vaccinations_by_age_ita %>%
  # remove cumulative sums, as they will be obtained via `summarise`:
  select(1:12) %>%
  group_by(fascia_anagrafica, fornitore) %>%
  summarise(across(where(is.numeric), sum)) %>%
  rename(vaccinati_totale = nuovi_vaccinati) -> totals_by_age_ita

totals_by_age_ita %>% head()
```

See how first and second shots are distributed by age range and supplier. Arguably, this is not very informative. But let's wait a couple of months: I bet it will prove useful!

```{r}
totals_by_age_ita %>%
  # we don't need all of these columns
  select(fascia_anagrafica, fornitore, prima_dose, seconda_dose) %>%
  # shift to long format and exclude cols that can't be pivoted
  pivot_longer(!c(fascia_anagrafica, fornitore),
               # give names to the new cols
               names_to = 'values',
               values_to = 'counts') %>%
  ggplot(aes(fornitore, counts, fill = values)) +
  geom_col() +
  facet_wrap(~ fascia_anagrafica) +
  scale_fill_viridis_d(begin = 0.55, end = 0.95) +
  theme_minimal() +
  labs(
    title = 'Number of shots by supplier, per age range',
    x = 'date',
    y = NULL,
    fill = NULL
  ) +
  theme(legend.position = 'bottom')
```

We can also picture the data as percentages:

```{r}
totals_by_age_ita %>%
  # we don't need all of these columns
  mutate(across(
    # across these two cols
    c(prima_dose, seconda_dose),
    # the new cols will be called {colname}.pct 
    list(pct = ~ .x / (prima_dose + seconda_dose))
  )) %>%
  select(fascia_anagrafica, fornitore, prima_dose_pct, seconda_dose_pct) %>%
  # shift to long format and exclude cols that can't be pivoted
  pivot_longer(!c(fascia_anagrafica, fornitore),
               # give names to the new cols
               names_to = 'values',
               values_to = 'counts') %>%
  ggplot(aes(fornitore, counts, fill = values)) +
  geom_col() +
  facet_wrap(~ fascia_anagrafica) +
  scale_fill_viridis_d(begin = 0.55, end = 0.95) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  labs(
    title = '% of shots by supplier, per age range',
    x = 'date',
    y = NULL,
    fill = NULL
  ) +
  theme(legend.position = 'bottom')
```

Does it really tell something? Well, it might not. But these posts are meant to document the process to find the better one and, of course, practice! We can do better using the time series data:

```{r}
vaccinations_by_age_ita %>%
  select(data, fascia_anagrafica, fornitore, prima_dose, seconda_dose) %>%
  pivot_longer(
    !c(data,fascia_anagrafica, fornitore),
    names_to = 'values',
    values_to = 'counts'
  ) %>%
  ggplot(aes(data, counts, fill = values)) +
  geom_col() +
  facet_wrap(~ fascia_anagrafica) +
  scale_fill_viridis_d(begin = 0.55, end = 0.95) +
  theme_minimal() +
  labs(
    title = 'Vaccine Shots by Age Range',
    x = 'date',
    y = NULL,
    fill = NULL
  ) +
  theme(legend.position = 'bottom')
```

This does not tell much either: the additional dimension of the doses is not enriched by these additional dimensions (but it's pretty good looking, is it not?).

# Group Data By Area

Let's wrap up by grouping the data by `area`.

```{r}
vaccinations_ita %>%
  group_by(data, area, fornitore) %>%
  # create new columns with the sum of the grouped rows
  summarise(across(where(is.numeric), sum)) %>%
  # reorder
  relocate(nuovi_vaccinati, .after = area) %>%
  # create new rows with cumulative sums
  mutate(across(where(is.numeric), list(totale = ~ cumsum(.x)))) %>%
  rename(vaccinati_totale = nuovi_vaccinati_totale) %>%
  arrange(area) -> vaccinations_by_area_ita
```

Before grouping again to get the totals, we want to enrich this data with the population and deliveries data. This script is a bit more elaborate:

```{r}
read_csv(
  'https://raw.githubusercontent.com/orizzontipolitici/covid19-vaccine-data/main/data_ita/doses_by_area_ita.csv'
) -> doses_by_area

vaccinations_by_area_ita %>%
  # `doses_by_area` is wide in format: we don't need `fornitore`
  select(1:12, -fornitore) %>%
  group_by(area) %>%
  summarise(across(where(is.numeric), sum)) %>%
  rename(vaccinati_totale = nuovi_vaccinati) %>%
  # also, it has the column for Italy, so we have to add one new line
  add_row(area = 'ITA',
          # the `.` indicates the object itself!
          vaccinati_totale = sum(.$vaccinati_totale),
          sesso_maschile = sum(.$sesso_maschile),
          sesso_femminile = sum(.$sesso_femminile),
          operatori_sanitari = sum(.$operatori_sanitari),
          personale_non_sanitario = sum(.$personale_non_sanitario),
          ospiti_rsa = sum(.$ospiti_rsa),
          over80 = sum(.$over80),
          prima_dose = sum(.$prima_dose),
          seconda_dose = sum(.$seconda_dose),
  ) %>%
  # perform inner join by area!
  inner_join(doses_by_area, by = 'area') %>%
  relocate(
    nome, NUTS2, area, popolazione_2020
  ) %>%
  # create new columns:
  mutate(
    # vaccinated each 1000 people
    vaccinati_ogni_mille = round(vaccinati_totale / popolazione_2020 * 1000, digits = 2),
    # share of vaccines used out of all received
    percent_vaccini_usati = round(vaccinati_totale / totale_dosi * 100, digits = 2)
  ) -> totals_by_area_ita

totals_by_area_ita
```