---
title: 'How Are Covid-19 Vaccinations Going On in Italy?'
author: 'Luca Baggi'
date: '2021-01-27'
slug: vis-covid-19-vaccine-italy
categories: []
tags:
  - covid19
ShowToc: true
---

This is just a very brief update to see how vaccinations are proceeding. I hope I can further develop this blogpost using `shiny` to make some basic visualisations interactive!

```{r, message=FALSE}
library(tidyverse)

theme_set(theme_minimal())
```

Let's load the data. We want to plot how many second doses are being administered each day.

```{r}
url <- 'https://raw.githubusercontent.com/orizzontipolitici/covid19-vaccine-data/main/data_ita/vaccinations_ita.csv'

vaccinations <- read_csv(url,
                         col_types = cols(
                           fornitore = col_factor(),
                           area = col_factor(),
                           fascia_anagrafica = col_factor()
                         )) %>%
  # drop over80 since it was not updated
  select(-over80)
```

What we want to do is `group_by()` the data by date and plot the cumulative distributions of `prima_dose` and `seconda_dose`.

Let's use `tidyr::complete()` in conjunction with `tidyr::full_seq()`: `full_seq(vaccinations$data, 1)` will fill the missing values, without needing to create the [grid](https://coding-corner.netlify.app/post/covid-19-vaccinations-italy-visualisations/) we used before.

```{r}
vaccinations %>%
  complete(data = full_seq(vaccinations$data, 1))
```
